---
title: "scrapeheartpdf"
author: "Rebecca Johnston"
date: "September 15, 2018"
output: html_document
---

Our aim is to scrape data from haematology pdf documents using R.

## Scraping pdfs

First attempt: scraping pdf using the R package `tm` which requires the external application [xpdfreader](http://www.xpdfreader.com/download.html) which is a bit of a pain to install for little gain (compared to the alternative package, `pdftools`, below).

```{r}
# install.packages("tm")
# library("tm")
# 
# files <- list.files(path = "data/", pattern = "pdf$")
# read <- readPDF(control = list(text = "-layout"))
# testdoc <- Corpus(URISource("data/B1705_2000560947.pdf"),
#                   readerControl = list(reader = read))
# doc <- content(testdoc[[1]])
# head(doc)
```

Second attempt: scraping pdf using the R package `pdftools`. I obtain similar results to `tm`, which is ultimately one big string you need to manipulate. Good help from [this](http://www.brodrigues.co/blog/2018-06-10-scraping_pdfs/) blog.

```{r}
library("pdftools")
library("tidyverse")

# Read in sample pdf
raw_table <- pdf_text("../sample-data/B1705_2000606271.pdf")

# Check structure
head(raw_table)
class(raw_table)
raw_table[[1]]

# Split character based on new line
split_table <- str_split(raw_table, "\r\n", simplify = TRUE)
```

```{r}
# Collect haematology data
table_start <- stringr::str_which(split_table, "RBC")
table_end <- stringr::str_which(split_table, "BLOOD SMEAR")
haem_table <- split_table[1, (table_start):(table_end)]
# Replace white space with underscore then split based on underscore 
haem_table <- str_replace_all(haem_table, "\\s{2,}", "_")
haem_table <- str_split_fixed(haem_table, "_", n = 4)
haem_table <- as.data.frame(haem_table)
# Keep only first two columns
haem_table <- haem_table %>% select(V1, V2)
```

```{r}
# Collect header information for metadata
meta_start <- stringr::str_which(split_table, "Animal ID")
meta_end <- stringr::str_which(split_table, "Reported")
meta_table <- split_table[1, (meta_start):(meta_end)]
# Replace white space with underscore then split based on underscore 
meta_table <- str_replace_all(meta_table, "\\s{2,}", "_")
# Only keep IDs we need within the string
meta_table[1] <- gsub("(^.*)(Animal ID.*$)", "\\2", meta_table[1])
meta_table[2] <- gsub("(^.*)(Unique ID.*$)", "\\2", meta_table[2])
# Add underscore between collected and received entries
meta_table[3] <- gsub("(^Collected.*)\\s(Received.*$)", "\\1_\\2", meta_table[3])
# Separate the strings
meta_table <- str_split(meta_table, "_", simplify = FALSE)
meta_table <- unlist(meta_table)
meta_table <- as.data.frame(meta_table)
meta_table <- str_split_fixed(meta_table$meta_table, ":\\s*", n = 2)
meta_table <- as.data.frame(meta_table)
```

```{r}
# Collect footer information, i.e. the "Examination" entry across multiple lines
foot_start <- stringr::str_which(split_table, "EXAMINATION")
foot_end <- stringr::str_which(split_table, "For tests indicated by a hash")
foot_table <- split_table[1, (foot_start):(foot_end - 1)]
foot_table <- str_replace_all(foot_table, "\\s{2,}", "")
# Separate "Examination" variable title
foot_table[1] <- gsub("(^EXAMINATION)(.*$)", "\\1_\\2", foot_table[1])
# Store the remainder as one long string
foot_table <- unlist(foot_table)
foot_table <- stringr::str_c(foot_table, sep = "", collapse = " ")
foot_table <- str_split(foot_table, "_", simplify = FALSE)
foot_table <- unlist(foot_table)
foot_table <- as.data.frame(matrix(foot_table, ncol = 2, byrow = TRUE))
```

So now we end up with three tables that we wish to join, `meta_table`, `haem_table`, and `foot_table`. The final step is to join them all together!
```{r}
bind_rows(meta_table, haem_table, foot_table)
```

## Function to clean pdf

Now we need to create a function to clean a pdf, that utilises all the above code, but performs everything in one step.
```{r}
clean_pdf <- function(raw_pdf){
  split_table <- str_split(raw_pdf, "\r\n", simplify = TRUE)
  
  # Collect haematology data
  table_start <- stringr::str_which(split_table, "RBC")
  table_end <- stringr::str_which(split_table, "BLOOD SMEAR")
  haem_table <- split_table[1, (table_start):(table_end)] %>% 
    str_replace_all(., "\\s{2,}", "_") %>%
    str_split_fixed(., "_", n = 4) %>%
    as.data.frame() %>%
    select(V1, V2)
  
  # Collect header information for metadata
  meta_start <- stringr::str_which(split_table, "Animal ID")
  meta_end <- stringr::str_which(split_table, "Reported")
  meta_table <- split_table[1, (meta_start):(meta_end)] %>%
    str_replace_all(., "\\s{2,}", "_")
  meta_table[1] <- gsub("(^.*)(Animal ID.*$)", "\\2", meta_table[1])
  meta_table[2] <- gsub("(^.*)(Unique ID.*$)", "\\2", meta_table[2])
  meta_table[3] <- gsub("(^Collected.*)\\s(Received.*$)", "\\1_\\2",
                        meta_table[3])
  meta_table <- str_split(meta_table, "_", simplify = FALSE) %>%
    unlist(.) %>%
    str_split_fixed(., ":\\s*", n = 2) %>%
    as.data.frame()

  # Collect footer information
  foot_start <- stringr::str_which(split_table, "EXAMINATION")
  foot_end <- stringr::str_which(split_table, "For tests indicated by a hash")
  foot_table <- split_table[1, (foot_start):(foot_end - 1)] %>%
    str_replace_all(., "\\s{2,}", "")
  foot_table[1] <- gsub("(^EXAMINATION)(.*$)", "\\1_\\2", foot_table[1])
  foot_table <- unlist(foot_table) %>%
    stringr::str_c(., sep = "", collapse = " ") %>%
    str_split(., "_", simplify = FALSE) %>%
    unlist(.)
  foot_table <- as.data.frame(matrix(foot_table, ncol = 2, byrow = TRUE))

  # Bind tables
  pdf_table <- bind_rows(meta_table, haem_table, foot_table)
}

# Test out the function
test <- clean_pdf(raw_table)
```

Now the REAL test is to try our code out on the other pdf files...
```{r}
raw_table_947 <- pdf_text("../sample-data/B1705_2000560947.pdf")
test_947 <- clean_pdf(raw_table_947)

raw_table_948 <- pdf_text("../sample-data/B1705_2000560948.pdf")
test_948 <- clean_pdf(raw_table_948)

raw_table_949 <- pdf_text("../sample-data/B1705_2000560949.pdf")
test_949 <- clean_pdf(raw_table_949)
```

